Computational methods in landscape ecology

### 2024-09-13 Sections/01-introduction.qmd
Minor cosmetic changes and starting to work on discussion

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Prlic2012TenSimple; @Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
Thus, future progress within the research field will highly depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].

<!-- MH: Given the limited word count I would cut this
For all computational methods the natural world must be translated into representative data models.
The spatial vector data model is based on points, lines, and polygons features that are defined by their geometry and additional attributes related to them.
Contrastingly, the raster data model is based on evenly-spaced grid cells and discrete or continuous values related to each cell [@Lovelace2019Geocomputation; @With2019EssentialsLandscape].
The choice of the data models, spatial and temporal scales, or the thematic resolution are crucial for the analysis [@Wade2003ComparisonVector; @Lechner2016RecentProgress; @Nedd2021SynthesisLand].
-->

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
Instead, for a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
<!-- JN: the first and the second point seem to be mixed in the text atm-->
Here, we first introduce recent computational developments in the most important topics of landscape ecology.
<!-- This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations. -->
Second, we present common software tools that implement or are potentially capable of implementing them.
Lastly, we discuss potential future directions for the development of computational methods in landscape ecology.

<!--JN: I am not the biggest fan of the following definition of open science: full transparency is rarely a goal IMO.-->
"Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields, such as landscape ecology, can massively benefit from this concept [@Hampton2015TaoOpen].
Using open source software is one key aspect of open science because it facilitates shareability, reproducibility, and transparency and consequently increases trust in the analyses [@Powers2019OpenScience; @Hesselbarth2021OpensourceTools].
<!-- In this context, script-based workflows have an advantage over workflows involving graphical user interfaces because the former are more shareable and reproducible [@Hampton2015TaoOpen; @Powers2019OpenScience]. -->
Thus, we will focus here on open-source software, specifically scripting languages such as R, Python, and Julia.

>>>>

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Prlic2012TenSimple; @Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
Instead, for a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing these developments 
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
Lastly, we discuss potential future directions for the development of computational methods in landscape ecology.

### 2024-09-13 Sections/02-spatial-patterns.qmd
Minor cosmetic changes and starting to work on discussion

Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpen].

Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].

@@

<!--JN: I would suggest to include the following paragraph, which I think is important to show the current state of the affairs.-->
The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
@Stepinski2022 and @Huckeba2023 showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
They argue that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
Thus, more research is needed to understand the relations between the entropy measures and how they relate to the environmental processes.

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies<!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRao], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

@@

R packages for such operations include *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource].
Popular Python libraries for spatial operations include *Geopandas* [@Jordahl2020GeopandasPython], *Shapely*[@Gillies2024ShapelyPython], or *PySal*[@Rey2010PySALPython].
The *GeoStats.jl* [@Hoffimann2018GeoStatsJl] and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] software provide general spatial data operations in the Julia programming language.

>>>>

Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpensource].

Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpensource], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].

@@

<!--JN: I would suggest to include the following paragraph, which I think is important to show the current state of the affairs.-->
The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
@Stepinski2022CurbYour and @Huckeba2023MultiscaleSpatial showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
They argue that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
Thus, more research is needed to understand the relations between the entropy measures and how they relate to the environmental processes.

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies<!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

@@

R packages for such operations include *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource].
Popular Python libraries for spatial operations include *Geopandas* [@Jordahl2020GeopandasPython], *Shapely*[@Gillies2024ShapelyPython], or *PySal*[@Rey2010PySALPython].
The *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] software provide general spatial data operations in the Julia programming language.

### 2024-09-13 Sections/03-connectivity.qmd
Minor cosmetic changes and starting to work on discussion

The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

<!--maybe add "Julia's", "R's", "Python's" before the software names?-->
Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].

>>>>

The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

<!--maybe add "Julia's", "R's", "Python's" before the software names?-->
Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapejlSoftware], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
